The non-random assembly of network motifs in  \nplant-pollinator networks

### 2022-10-08 Manuscript/Main_text/Main_text_Journal_of_Animal_Ecol.Rmd
Minor edits

---
title: "The non-random assembly of indirect interactions in  \nplant-pollinator networks"
editor_options:
  chunk_output_type: console
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
- \usepackage{caption,setspace}
- \captionsetup[figure]{labelformat=empty}
- \floatplacement{figure}{H}
- \usepackage[skip=3pt]{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{lscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage[left]{lineno}
- \linenumbers
output:
  pdf_document: default
  html_document: default
  word_document: default
always_allow_html: true
bibliography: bibliography.bib
biblio-style: apalike
csl: JournalOfAnimalEcology.csl #converted from funct ecol style from zotero styles
---

@@

Ecological processes leave distinct structural imprints on the species interactions that shape the topology of animal-plant mutualistic networks. Detecting how direct and indirect interactions between animals and plants are organised is not trivial since they go beyond pairwise interactions, but may get blurred when considering global network descriptors. Recent work has shown that the meso-scale, the intermediate level of network complexity between the species and the global network, can capture this important information. The meso-scale describes network subgraphs representing patterns of direct and indirect interactions between a small number of species and when these network subgraphs differ statistically from a benchmark, they are often referred to as "network motifs". Although motifs can capture relevant ecological information of species interactions, they remain overlooked in natural plant-pollinator networks. By exploring 60 empirical plant-pollinator networks from 18 different studies with wide geographical coverage we show that some network subgraphs are consistently under- or over-represented, suggesting the presence of worldwide network motifs in plant-pollinator networks. In addition, we found a higher proportion of densely connected network subgraphs that, based on previous findings, could reflect that species relative abundances are the main driver shaping the structure of the meso-scale on plant-pollinator communities. Moreover, we found that distinct subgraph positions describing species ecological roles (e.g., generalisation and number of indirect interactions) are occupied by different groups of animal and plant species representing their main life-history strategies (i.e., functional groups). For instance, we found that the functional group of 'bees' was over-represented in subgraph positions with a lower number of indirect interactions in contrast to the rest of floral visitors groups. Finally, we show that the observed functional group combinations within a subgraph can not be retrieved from their expected probabilities (i.e., joint probability distributions), indicating that plant and floral visitor associations within subgraphs are not random either. Therefore, our results highlight the presence of common motifs in plant-pollinator communities that are formed by a non-random association of plants and floral visitors functional groups.

@@

Indirect interactions, the mediated effect between two species by a third species, are ubiquitous in natural systems [@strauss1991; @holt1994; @wootton1994] and play a key role in shaping community dynamics [@werner2003; @walsh2013]. Although some of the most studied indirect interactions are the ones caused by trophic cascades with consumer-resource interactions [@abrams1996; @schmitz2001], previous studies often focus on antagonistic interactions and overlook mutualistic ones [@knight2005]. On the contrary, plant-pollinator research has traditionally investigated direct interactions but neglected indirect effects such as competitive or facilitative processes between plants for pollinators [@moeller2004; @sargent2008; @carvalheiro2014]. Conventional analytical tools in bipartite networks often discard relevant information about indirect interactions by condensing the information into single indices either at species or network level [@cirtwill2018; @simmons2019]. However, as previously explored with three species modules in unipartite food webs [@allesina2008; @stouffer2010], the emergent framework of bipartite subgraphs has proven to be a powerful analytical tool to better understand plant-pollinator community dynamics by capturing both direct and indirect effects [@simmons2019]. These network subgraphs that constitute the meso-scale are often referred to as the building blocks of the network and when their numbers occur at higher frequencies than in randomised networks, those are generally defined as 'network motifs' [@milo2002]. Indeed, the analysis of motifs in simulated plant-pollinator networks has revealed that the different drivers that govern species interactions (e.g., species abundances versus trait-matching) can lead to different patterns of indirect effects [@simmons2020]. However, the global patterns of indirect interactions in real plant-pollinator networks and their ecological consequences remain unexplored (e.g., over- and under- represented network subgraphs).

@@

First, plant species were grouped through hierarchical clustering into the optimal number of functional groups that summarises the main plant reproductive strategies. For this, we used the plant trait dataset compiled in @lanuza2021 that consisted also of plant species that belonged to the same plant-pollinator networks plus some additional meta-networks not considered here. This dataset comprised a total of 1506 plant species including the 503 species considered in this study (Table S2). Because most of these plant species are likely to benefit from animal pollination, we used reproductive traits relevant to understand the different plant reproductive strategies and their interactions with pollinators. After excluding traits with a high percentage of missing values (over 30$\%$), we ended up with 8 floral, 3 reproductive biology and 3 plant morphology traits (see Table S3 for a list of all traits). Floral traits included those associated with the size, form and number of floral organs (e.g., style length or ovule number). Reproductive biology traits described the reproductive system of the plant (e.g., mating and compatibility systems) and plant morphology traits were associated with the size and form of the plant (i.e., plant height, lifespan and life form). After an extensive search for each plant species on several different resources (e.g., plant databases, online floras, books, journals and images), over 80% of the cells of the whole dataset were filled. Thus, to avoid the loss of species with trait missing values, missing trait data was imputed as detailed in @lanuza2021.

@@

Following previous work [@simmons2019; @simmons2020], we broke down the plant-pollinator networks into their constituent network subgraphs that can capture both direct and indirect interactions. Prior to analyses, we turned the quantitative networks into qualitative (or binary) ones, where interactions are present or absent. This approach was preferred since quantitative subgraph analyses have been developed just recently and more work is needed to fully understand and interpret the information that they can convey [@simmons2020]. All analyses were run considering singletons (species with only one interaction detected) but we conducted a second exploration without singletons (64.98$\%$ of interactions) to evaluate the effect of rare species on the observed patterns. Results without singletons are qualitatively consistent and are not further discussed in the main text (Figs S2 and S3).

We calculated the frequency of all subgraphs up to five nodes (17 different subgraphs in total; see Fig. 1) for each empirical network, by using the *bmotif* package version *2.0.2* [@bmotifsimmons2019]. We focused exclusively on subgraphs up to five nodes in our analyses for computational reasons and for the ease of visualization and interpretation of a reduced set of less complex network structures. Throughout the entire manuscript, floral visitors occupy the top nodes of the different network subgraphs and plants the bottom ones. To control for variation in network size, subgraph frequencies were normalised as a proportion of the total number of subgraphs within each subgraph class (i.e., 2-species, 3-species, 4-species and 5-species network subgraphs). In addition, we excluded 2-species (or nodes) subgraphs from our analyses because its normalised frequencies would always equal one. Note that the aim is not comparing different networks, which may differ in sampling effort, but correctly describing the structure of each individual network. Like @simmons2020, we classified the different subgraphs by their average path length (mean number of links between all pairs of nodes) into four different groups: (i) complete or strong ($\overline{x} = 1.38$), (ii) fan or medium-strong ($\overline{x} = 1.48$), (iii) assymmetric complete or medium-weak ($\overline{x} = 1.60$) and (iv) core-peripheral or weak ($\overline{x} = 1.85$), where $\overline{x}$ represents the mean of the average path length across subgraphs within the same group. In general terms, plant and floral visitor groups on more densely connected network subgraphs (e.g., complete or fan) will experience lower number of indirect interactions in contrast to groups on less densely connected structures (e.g., medium-weak or weak).

@@

We calculated which plant and floral visitor functional groups were over- or under-represented in the different subgraph positions by comparing position frequencies of empirical networks with those of their corresponding simulated counterparts. To estimate the position frequencies of each plant and floral visitor group in a given network, we added the frequencies of those species that belong to the same group, and then, we normalised the resulting frequencies by dividing them by the total number of times that a corresponding group appears in any position within the same subgraph size class (i.e., 2, 3, 4 or 5 species motifs). Then, we calculated the percentile of the observed position frequencies for each functional group and network, just like we did with subgraph frequencies. To outline the general patterns of position frequencies across networks and groups, we fit a LMM per subgraph position, where the response variable was the observed position percentile per network. We used the group identifier as an explanatory variable and the study identifiers as a random intercept. Hence, we assessed the average subgraph frequency per species group, after controlling for the variation at the study level. Finally, we visualised over- and under-representation of plant and floral visitor groups on the different subgraph positions with the help of the *ComplexHeatmap* package version *2.6.2* [@gu2016].

To understand if the different plant and floral visitor functional groups are associated with subgraph positions that represent different network roles, we characterised the roles of each subgraph position for the different functional groups in terms of their specialisation and number of indirect interactions. To that end, we calculated the specificity of each subgraph position following @poisot2015 as $s_i = (L - l_i) / (L - 1)$, where $L$ is the total number of groups and $l_i$ the number of interaction partners of the group $i$ in a given subgraph. Values of the vector $si$, range between 1, complete specialisation (1 partner within the subgraph) and 0 complete generalism (all possible partners within the subgraph). Then, because all nodes within a motif are connected, we calculated the number of indirect interactions ($\theta_i$) for each motif position as $\theta_i=m_i-k_i-1$, where $m_i$ is the total number of nodes in a given motif, and $k_i$ represents the degree of the corresponding motif position, that is, the total number of links connected to such position. Thus, we considered all possible indirect interactions for each position. Finally, we analysed in a single model how specialisation and the number of indirect interactions were associated with the observed percentiles for each trophic guild and subgraph position (see paragraph above). For this, we fitted again a LMM model with the help of the *lme4* package where we considered specialisation and number of indirect interactions as response variables with the interaction by plant and floral visitor group for both response variables. Further, we considered as random factor subgraph position nested within study system.

@@

The hierarchical cluster analysis divided the dataset into five different plant functional groups with different and overlapping characteristics (Figs S6 and S7). The first group referred to as 'selfing herbs' consisted of herbs with hermaphroditic flowers with high levels of autonomous selfing. The second group named 'small outcrossing perennials' had small perennial species with a mix of life forms (i.e., trees, shrubs and herbs) with outcrossing hermaphroditic flowers. The third group referred to as 'self-incompatible perennials with large flowers' comprised perennial species with a mixed of life forms and large self-incompatible hermaphroditic flowers with high number of ovules. The fourth group named 'tall plants with small unisexual flowers' had the tallest species, highest proportion of shrub and tree life forms, dioecious and monoecious breeding systems, small flowers and the highest numbers of flowers and inflorescences per plant. Finally, the fifth group named 'short-lived outcrossers with long zygomorphic flowers' consisted of small perennial and short-lived herbs with long self-compatible zygomorphic flowers that were unable to self-pollinate.

@@

The comparison of the plant and floral visitor functional group frequencies per subgraph position between empirical and simulated networks showed a `r over`$\%$ and `r under`$\%$ of over- (>75$^{th}$) and under-represented (\<25th) functional groups in the different positions, respectively (Fig. 3). Floral visitors showed a total of `r poll_over_75`$\%$ and `r poll_under_25`$\%$ of over- and under-represented functional groups in the different positions and plants `r plant_over_75`$\%$ and `r plant_under_25`$\%$ of over- and under-represented functional groups, respectively. Notably, the differences across functional groups were more marked for floral visitors than for plants (the differences between min and max percentiles per position were generally two-three times larger for floral visitors). From most over- to under-represented floral visitors functional groups on the different subgraph positions (indicated with the dendrogram order in Fig. 3), we found: bees, non-syrphids Diptera, syrphids, Coleoptera, non-bee Hymenoptera and Lepidoptera. Although plant functional groups showed less differences between them, there were also more represented functional groups than others, thus from most over- to under-represented functional groups on the different subgraph positions we found: self incompatible perennials with large flowers, small outcrossing perennials, tall plants with unisexual flowers, selfing herbs and short lived outcrossers with long zygomorphic flowers.

For both floral visitor and plant functional groups we found over- and under-representation on different subgraph positions that implied different ecological roles. First, for floral visitors, we found that bees were less frequent on subgraph positions that involved specialisation (i.e., lower specificity values) while the rest of functional groups (non-syrphid Diptera, syrphids, Lepidoptera, Coleoptera and non-bee Hymenoptera) were infrequent in generalised subgraph positions with low number of indirect interactions (Fig. S8 A). Although plants did not show differences with null models on their frequencies, all plant functional groups but tall plants with small unisexual flowers showed a slight tendency to be over-represented on specialised subgraph positions (Fig. S8 B). Second, regarding the number of indirect interactions, bees were over-represented in subgraph positions with low number of indirect interactions (Fig. S9 A) and the rest of floral visitor functional groups showed no differences across positions in comparison with simulated networks. For plants, all functional groups showed a slight tendency to be over-represented on subgraph positions with higher number of indirect interactions (Fig. S9 B). When we further explore the observed frequencies by number of indirect interactions for floral visitor and plant functional groups, we found that both bees and tall plants with small unisexual flowers were the only functional groups with high frequencies on subgraph positions associated with low number interactions in comparison with the other floral visitor and plant functional groups (Figs S10 and S11).

@@

The statistical comparison with z-scores between the observed and expected plant and floral visitor functional group combinations showed that we can not recover the observed combinations from its probability of occurrence. In fact, most combinations are under- or over- represented and follow a modified Gaussian distribution (Johnson's $S_U$) with 65$\%$ of under-represented group combinations, 20$\%$ of no statistical difference and 15$\%$ of under-represented ones (Fig. 5). In addition, subgraphs with small node combinations (i.e., 2 and 3 nodes) only appeared as under-represented. Network subgraphs with 4 and 5 node combinations appeared in the three statistical categories but 4 node subgraphs had a higher proportion in the under-represented category (10% in comparison with the 1$\%$ and the 3$\%$ of the no statistical difference and over-represented categories, respectively). Finally, 5 node subgraphs were in similar proportions in the no statistical difference and over-represented categories (99$\%$ and 97$\%$, respectively) but they appeared in lower proportion in the under-represented category (89$\%$). When we explored the identity on the different positions of the most probable subgraph combinations we found that the most common plant and floral visitor functional groups were also the most probable on these plant and floral visitor combinations within the different subgraphs (lower panel Fig. 5).

@@

Understanding the global patterns of network structure is key to unravel the processes that govern community dynamics [@bascompte2007; @bastolla2009; @guimaraes2020]. However, common analytical approaches at the node or at the global network level are unable to capture the full complexity of ecological systems [@simmons2019; @bergamo2021]. Because network subgraphs can capture both direct and indirect interactions, by analysing their over- and under-representation we were able to explore in great detail the structure of interactions between animal pollinators and the plants they visit. We have explored for the first time the overall structure and presence of network motifs in real plant-pollinator networks and shown that the different subgraphs up to five nodes are consistently over- or under-represented in comparison with simulated networks. In addition, we show that for plant-pollinator networks there is a general tendency for over-representation of network subgraphs that are more densely connected (i.e., complete, fan and asymmetric complete). Indeed, the work of @simmons2020 has shown with the help of simulated plant-pollinator networks that the main different processes that shape plant-pollinator communities (i.e., neutral effects, trait matching and phenology) generate different patterns of direct and indirect interactions, and that specifically, a higher proportion of these more densely connected subgraphs is associated with neutral effects. Therefore, as indicated for other network properties like nestedness at the macro-scale level [@krishna2008; @suweis2013], or interaction probability at the species level [@bartomeus2016], our findings indicate that neutral effects may be also the main driver governing this intermediate level of complexity on plant-pollinator networks.

Overall, generalist network motifs for floral visitors were consistently over-represented (e.g., complete subgraphs; Fig. 2). For instance, when we compare network subgraphs with equal structure for both trophic levels, we found that floral visitors had always over-representation in the positions that implied the highest number of direct links (e.g., positions 16 and 46 from network subgraphs 7 and 17, respectively). Despite this being in good agreement with the current view of the generalist nature of pollinators [@waser1996; @olesen2002; @ollerton2017], we found that this over-representation of generalist network motifs was driven by the functional group of bees while the rest of the functional groups tended to be under-represented (Fig. S6 A). In contrast, non-bee groups were over-represented in subgraph positions associated with specialised roles. This large effect of bees on the overall patterns is no surprise, as bees were the group with the largest number of interaction records in the whole set of networks as also shown for other broad scale plant-pollinator network studies [@carvalheiro2014; @ollerton2017]. Just syrphids had a relatively similar number of interactions in comparison with bees (1,768 and 2,256 respectively) and similar species richness (343 and 321, respectively). However, both functional groups were associated with contrasting ecological roles likely indicating that both number of interactions and richness are not driving the observed interaction patterns for the different functional groups. Interestingly, bees also were the group with lower number of indirect interactions while the rest of floral visitor groups showed a clearly higher number of indirect interactions indicating that these non-bee floral visitor groups may experience higher competitive interactions for floral resources [@thomson2020]. Despite plant functional groups showed little differences between them on their subgraph frequencies, all groups showed a tendency to be over-represented on subgraph positions that were associated with higher levels of specialisation. This tendency towards specialist roles may be the result of co-evolutionary processes (e.g., trait-matching) between plants and pollinators to optimise pollen transfer [@moreira2019]. Remarkably, only tall plants with small unisexual flowers showed higher frequencies on subgraphs that were associated with low number of indirect interactions with likely direct implications on pollen transfer for this group that needs to guarantee cross-pollination.

>>>>

---
title: "The non-random assembly of network motifs in  \nplant-pollinator networks"
editor_options:
  chunk_output_type: console
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
- \usepackage{caption,setspace}
- \captionsetup[figure]{labelformat=empty}
- \floatplacement{figure}{H}
- \usepackage[skip=3pt]{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{lscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage[left]{lineno}
- \linenumbers
output:
  pdf_document: default
  html_document: default
  word_document: default
always_allow_html: true
bibliography: bibliography.bib
biblio-style: apalike
csl: JournalOfAnimalEcology.csl #converted from funct ecol style from zotero styles
---

@@

Ecological processes leave distinct structural imprints on the species interactions that shape the topology of animal-plant mutualistic networks. Detecting how direct and indirect interactions between animals and plants are organised is not trivial since they go beyond pairwise interactions, but may get blurred when considering global network descriptors. Recent work has shown that the meso-scale, the intermediate level of network complexity between the species and the global network, can capture this important information. The meso-scale describes network subgraphs representing patterns of direct and indirect interactions between a small number of species, and when these network subgraphs differ statistically from a benchmark, they are often referred to as "network motifs". Although motifs can capture relevant ecological information of species interactions, they remain overlooked in natural plant-pollinator networks. By exploring 60 empirical plant-pollinator networks from 18 different studies with wide geographical coverage, we show that some network subgraphs are consistently under- or over-represented, suggesting the presence of worldwide network motifs in plant-pollinator networks. In addition, we found a higher proportion of densely connected network subgraphs that, based on previous findings, could reflect that species relative abundances are the main driver shaping the structure of the meso-scale on plant-pollinator communities. Moreover, we found that distinct subgraph positions describing species ecological roles (e.g., generalisation and number of indirect interactions) are occupied by different groups of animal and plant species representing their main life-history strategies (i.e., functional groups). For instance, we found that the functional group of 'bees' was over-represented in subgraph positions with a lower number of indirect interactions in contrast to the rest of floral visitors groups. Finally, we show that the observed functional group combinations within a subgraph can not be retrieved from their expected probabilities (i.e., joint probability distributions), indicating that plant and floral visitor associations within subgraphs are not random either. Therefore, our results highlight the presence of common network motifs in plant-pollinator communities that are formed by a non-random association of plants and floral visitors functional groups.

@@

Indirect interactions, the mediated effect between two species by a third species, are ubiquitous in natural systems [@strauss1991; @holt1994; @wootton1994] and play a key role in shaping community dynamics [@werner2003; @walsh2013]. Some of the most studied indirect interactions are the ones caused by trophic cascades with consumer-resource interactions [@abrams1996; @schmitz2001]. In fact, previous studies focus on antagonistic interactions and mutualistic interactions are often overlooked [@knight2005]. On the contrary, plant-pollinator research has traditionally investigated direct interactions but neglected indirect effects such as competitive or facilitative processes between plants for pollinators [@moeller2004; @sargent2008; @carvalheiro2014]. Conventional analytical tools in bipartite networks often discard relevant information about indirect interactions by condensing the information into single indices either at species or network level [@cirtwill2018; @simmons2019]. However, as previously explored with three species modules in unipartite food webs [@allesina2008; @stouffer2010], the emergent framework of bipartite subgraphs has proven to be a powerful analytical tool to better understand plant-pollinator community dynamics by capturing both direct and indirect effects [@simmons2019]. These network subgraphs that constitute the meso-scale are often referred to as the building blocks of the network and when their numbers occur at higher frequencies than in randomised networks, those are generally defined as 'network motifs' [@milo2002]. Indeed, the analysis of motifs in simulated plant-pollinator networks has revealed that the different drivers that govern species interactions (e.g., species abundances versus trait-matching) can lead to different patterns of indirect effects [@simmons2020]. However, the global patterns of indirect interactions in real plant-pollinator networks and their ecological consequences remain unexplored (e.g., over- and under- represented network subgraphs).

@@

First, plant species were grouped through hierarchical clustering into the optimal number of functional groups that summarises the main plant reproductive strategies. For this, we used the plant trait dataset compiled in @lanuza2021 that consisted also of plant species that belonged to the same plant-pollinator networks plus some additional meta-networks not considered here. This dataset comprised a total of 1,506 plant species including the 503 species considered in this study (Table S2). Because most of these plant species are likely to benefit from animal pollination, we used reproductive traits relevant to understand the different plant reproductive strategies and their interactions with pollinators. After excluding traits with a high percentage of missing values (over 30$\%$), we ended up with 8 floral, 3 reproductive biology and 3 plant morphology traits (see Table S3 for a list of all traits). Floral traits included those associated with the size, form and number of floral organs (e.g., style length or ovule number). Reproductive biology traits described the reproductive system of the plant (e.g., mating and compatibility systems) and plant morphology traits were associated with the size and form of the plant (i.e., plant height, lifespan and life form). After an extensive search for each plant species on several different resources (e.g., plant databases, online floras, books, journals and images), over 80% of the cells of the whole dataset were filled. Thus, to avoid the loss of species with trait missing values, missing trait data was imputed as detailed in @lanuza2021.

@@

Following previous work [@simmons2019; @simmons2020], we broke down the plant-pollinator networks into their constituent network subgraphs that can capture both direct and indirect interactions. Prior to analyses, we turned the quantitative networks into qualitative (or binary) ones, where interactions are present or absent. This approach was preferred since quantitative subgraph analyses have been developed just recently and more work is needed to fully understand and interpret the information that they can convey [@simmons2020]. All analyses were run considering singletons (species with only one interaction detected) but we conducted a second exploration without singletons (64.98$\%$ of interactions) to evaluate the effect of rare species on the observed patterns. Results without singletons are qualitatively consistent and are not further discussed in the main text (Figs. S2 and S3).

We calculated the frequency of all subgraphs up to five nodes (17 different subgraphs in total; see Fig. 1) for each empirical network, by using the *bmotif* package version *2.0.2* [@bmotifsimmons2019]. We focused exclusively on subgraphs up to five nodes in our analyses for computational reasons and for the ease of visualization and interpretation of a reduced set of less complex network structures. Throughout the entire manuscript, floral visitors occupy the top nodes of the different network subgraphs and plants the bottom ones. To control for variation in network size, subgraph frequencies were normalised as a proportion of the total number of subgraphs within each subgraph class (i.e., 2-species, 3-species, 4-species and 5-species network subgraphs). In addition, we excluded 2-species (or 2-nodes) subgraphs from our analyses because its normalised frequencies would always equal one. Note that the aim is not comparing different networks, which may differ in sampling effort, but correctly describing the structure of each individual network. Like @simmons2020, we classified the different subgraphs by their average path length (mean number of links between all pairs of nodes) into four different groups: (i) complete or strong ($\overline{x} = 1.38$), (ii) fan or medium-strong ($\overline{x} = 1.48$), (iii) assymmetric complete or medium-weak ($\overline{x} = 1.60$) and (iv) core-peripheral or weak ($\overline{x} = 1.85$), where $\overline{x}$ represents the mean of the average path length across subgraphs within the same group. In general terms, plant and floral visitor groups on more densely connected network subgraphs (e.g., complete or fan) will experience lower number of indirect interactions in contrast to groups on less densely connected structures (e.g., medium-weak or weak).

@@

We calculated which plant and floral visitor functional groups were over- or under-represented in the different subgraph positions by comparing position frequencies of empirical networks with those of their corresponding simulated counterparts. To estimate the position frequencies of each plant and floral visitor group in a given network, we added the frequencies of those species that belong to the same group, and then, we normalised the resulting frequencies by dividing them by the total number of times that a corresponding group appears in any position within the same subgraph size class. Then, we calculated the percentile of the observed position frequencies for each functional group and network, just like we did with subgraph frequencies. To outline the general patterns of position frequencies across networks and groups, we fit a LMM per subgraph position, where the response variable was the observed position percentile per network. We used the group identifier as an explanatory variable and the study identifiers as a random intercept. Hence, we assessed the average subgraph frequency per species group, after controlling for the variation at the study level. Finally, we visualised over- and under-representation of plant and floral visitor groups on the different subgraph positions with the help of the *ComplexHeatmap* package version *2.6.2* [@gu2016].

To understand if the different plant and floral visitor functional groups are associated with subgraph positions that represent different network roles, we characterised the roles of each subgraph position for the different functional groups in terms of their specialisation and number of indirect interactions. To that end, we calculated the specificity of each subgraph position following @poisot2015 as $s_i = (L - l_i) / (L - 1)$, where $L$ is the total number of groups and $l_i$ the number of interaction partners of the group $i$ in a given subgraph. Values of the vector $si$, range between 1, complete specialisation (1 partner within the subgraph) and 0 complete generalism (all possible partners within the subgraph). Then, because all nodes within a motif are connected, we calculated the number of indirect interactions ($\theta_i$) for each motif position as $\theta_i=m_i-k_i-1$, where $m_i$ is the total number of nodes in a given motif, and $k_i$ represents the degree of the corresponding motif position, that is, the total number of links connected to such position. Thus, we considered all possible indirect interactions for each position. Finally, we analysed in a single model how specialisation and the number of indirect interactions were associated with the observed percentiles for each trophic guild and subgraph position. For this, we fitted again a LMM model with the help of the *lme4* package, where we considered specialisation and number of indirect interactions as response variables with the interaction by plant and floral visitor group for both response variables. Further, we considered as random factor subgraph position nested within study system.

@@

The hierarchical cluster analysis divided the dataset into five different plant functional groups with different and overlapping characteristics (Figs. S4 and S5). The first group referred to as 'selfing herbs' consisted of herbs with hermaphroditic flowers with high levels of autonomous selfing. The second group named 'small outcrossing perennials' had small perennial species with a mix of life forms (i.e., trees, shrubs and herbs) with outcrossing hermaphroditic flowers. The third group referred to as 'self-incompatible perennials with large flowers' comprised perennial species with a mixed of life forms and large self-incompatible hermaphroditic flowers with high number of ovules. The fourth group named 'tall plants with small unisexual flowers' had the tallest species, highest proportion of shrub and tree life forms, dioecious and monoecious breeding systems, small flowers, and the highest numbers of flowers and inflorescences per plant. Finally, the fifth group named 'short-lived outcrossers with long zygomorphic flowers' consisted of small perennial and short-lived herbs with long self-compatible zygomorphic flowers that were unable to self-pollinate.

@@

The comparison of the plant and floral visitor functional group frequencies per subgraph position between empirical and simulated networks showed a `r over`$\%$ and `r under`$\%$ of over- (>75$^{th}$) and under-represented (\<25th) functional groups in the different positions, respectively (Fig. 3). Floral visitors showed a total of `r poll_over_75`$\%$ and `r poll_under_25`$\%$ of over- and under-represented functional groups in the different positions, and plants `r plant_over_75`$\%$ and `r plant_under_25`$\%$ of over- and under-represented functional groups, respectively. Notably, the differences across functional groups were more marked for floral visitors than for plants (the differences between the minimum and maximum values of the percentiles per position were generally two-three times larger for floral visitors). From most over- to under-represented floral visitors functional groups on the different subgraph positions (indicated with the dendrogram order in Fig. 3), we found: bees, non-syrphids Diptera, syrphids, Coleoptera, non-bee Hymenoptera and Lepidoptera. Although plant functional groups showed less differences between them, there were also more represented functional groups than others. Thus, from most over- to under-represented functional groups on the different subgraph positions we found: self incompatible perennials with large flowers, small outcrossing perennials, tall plants with unisexual flowers, selfing herbs and short lived outcrossers with long zygomorphic flowers.

For both floral visitor and plant functional groups we found over- and under-representation on different subgraph positions that implied different ecological roles. First, for floral visitors, we found that bees were less frequent on subgraph positions that involved specialisation (i.e., lower specificity values) while the rest of functional groups (non-syrphid Diptera, syrphids, Lepidoptera, Coleoptera and non-bee Hymenoptera) were infrequent in generalised subgraph positions with low number of indirect interactions (Fig. S6 A). Although plants did not show differences with null models on their frequencies, all plant functional groups but tall plants with small unisexual flowers showed a slight tendency to be over-represented on specialised subgraph positions (Fig. S6 B). Second, regarding the number of indirect interactions, bees were over-represented in subgraph positions with low number of indirect interactions (Fig. S7 A) and the rest of floral visitor functional groups showed no differences across positions in comparison with simulated networks. For plants, all functional groups showed a slight tendency to be over-represented on subgraph positions with higher number of indirect interactions (Fig. S7 B). When we further explore the observed frequencies by number of indirect interactions for floral visitor and plant functional groups, we found that both bees and tall plants with small unisexual flowers were the only functional groups with high frequencies on subgraph positions associated with low number of indirect interactions in comparison with the other floral visitor and plant functional groups (Figs. S8 and S9).

@@

The statistical comparison with z-scores between the observed and expected plant and floral visitor functional group combinations showed that we can not recover the observed combinations from its probability of occurrence. In fact, most combinations are under- or over- represented and follow a modified Gaussian distribution (Johnson's $S_U$) with 65$\%$ of under-represented group combinations, 20$\%$ of no statistical difference and 15$\%$ of under-represented ones (Fig. 5). In addition, subgraphs with small node combinations (i.e., 2 and 3 nodes) only appeared as under-represented. Network subgraphs with 4 and 5 node combinations appeared in the three statistical categories, but 4 node subgraphs had a higher proportion in the under-represented category (10% in comparison with the 1$\%$ and the 3$\%$ of the no statistical difference and over-represented categories, respectively). Finally, 5 node subgraphs were in similar proportions in the no statistical difference and over-represented categories (99$\%$ and 97$\%$, respectively), but they appeared in lower proportion in the under-represented category (89$\%$). When we explored the identity on the different positions of the most probable subgraph combinations, we found that the most common plant and floral visitor functional groups were also the most probable on these plant and floral visitor combinations within the different subgraphs (lower panel Fig. 5).

@@

Understanding the global patterns of network structure is key to unravel the processes that govern community dynamics [@bascompte2007; @bastolla2009; @guimaraes2020]. However, common analytical approaches at the node or at the global network level are unable to capture the full complexity of ecological systems [@simmons2019; @bergamo2021]. Because network subgraphs can capture both direct and indirect interactions, by analysing their over- and under-representation we were able to explore in great detail the structure of interactions between animal pollinators and the plants they visit. We have explored for the first time the overall structure and presence of network motifs in real plant-pollinator networks and shown that, according to our first hypothesis, the different subgraphs up to five nodes are consistently over- or under-represented in comparison with simulated networks. In addition, we show that for plant-pollinator networks there is a general tendency for over-representation of network subgraphs that are more densely connected (i.e., complete, fan and asymmetric complete). Indeed, the work of @simmons2020 has shown with the help of simulated plant-pollinator networks that the main different processes that shape plant-pollinator communities (i.e., neutral effects, trait matching and phenology) generate different patterns of direct and indirect interactions, and that specifically, a higher proportion of these more densely connected subgraphs is associated with neutral effects. Therefore, as indicated for other network properties like nestedness at the macro-scale level [@krishna2008; @suweis2013], or interaction probability at the species level [@bartomeus2016], our findings align with our second hypothesis and indicate that neutral effects may be also the main driver governing this intermediate level of complexity on plant-pollinator networks.

Overall, generalist network motifs for floral visitors were consistently over-represented (e.g., complete subgraphs; Fig. 2). For instance, when we compare network subgraphs with equal structure for both trophic levels, we found that floral visitors had always over-representation in the positions that implied the highest number of direct links (e.g., positions 16 and 46 from network subgraphs 7 and 17, respectively). Despite this being in good agreement with the current view of the generalist nature of pollinators [@waser1996; @olesen2002; @ollerton2017], we found that this over-representation of generalist network motifs was driven by the functional group of bees while the rest of the functional groups tended to be under-represented (Fig. S6 A). In contrast, non-bee groups were over-represented in subgraph positions associated with specialised roles. This large effect of bees on the overall patterns is no surprise, as bees were the group with the largest number of interaction records in the whole set of networks as also shown for other broad scale plant-pollinator network studies [@carvalheiro2014; @ollerton2017]. Just syrphids had a relatively similar number of interactions in comparison with bees (1,768 and 2,256 interactions, respectively) and similar species richness (343 and 321 species, respectively). However, both functional groups were associated with contrasting ecological roles likely indicating that both number of interactions and richness are not driving the observed interaction patterns for the different functional groups. Interestingly, bees also were the group with lower number of indirect interactions, while the rest of floral visitor groups showed a clearly higher number of indirect interactions, indicating that these non-bee floral visitor groups may experience higher competitive interactions for floral resources [@thomson2020]. Despite plant functional groups showed little differences between them on their subgraph frequencies, all groups showed a tendency to be over-represented on subgraph positions that were associated with higher levels of specialisation. This tendency towards specialist roles may be the result of co-evolutionary processes (e.g., trait-matching) between plants and pollinators to optimise pollen transfer [@moreira2019]. Remarkably, only tall plants with small unisexual flowers showed higher frequencies on subgraphs that were associated with low number of indirect interactions with likely direct implications on pollen transfer for this group that needs to guarantee cross-pollination.

The probabilities of appearance of each functional group on the different subgraph positions were insufficient to predict most of the observed network position combinations (Fig. 5). This implies that not just the representation of network subgraphs types but also the identity of the elements of their structure is non-random. Notably, we found that just 30$\%$ of the possible 53,250 subgraph combinations are realised. Thus, similarly to the presence of forbidden links at the network level [@olesen2011; @jordano2016], we also found the existence of forbidden subgraph positions and those represented almost two thirds of the total possible subgraph combinations. This concurs well with the general idea that plant-pollinator networks have a high proportion of non-realized direct interactions [@jordano2003; @bascompte2013], and, as a result, also a high proportion of non-realized combinations of indirect interactions. Some of these non-observed direct interactions are probably due to under-sampling [@dorado2011; @jordano2016]. However, a large fraction of the possible interactions do not take place because of different constraints such as species co-occurrence, trait matching processes and phenological uncoupling [@olesen2011; @bartomeus2016; @jordano2016]. Despite the advances in the understanding of the ecological processes that influence species pairwise interactions [@bartomeus2016; @peralta2020], we still know little about how these processes determine competitive and facilitative interactions in mutualistic networks. Thus, as suggested by @simmons2020, exploring the proportion of these realised network subgraphs that are competitive or facilitative is a necessary next step to improve our understanding of the architecture of biodiversity [@bascompte2007].
